---
title: 'Final Project'
subtitle: "PSTAT122: Design and Analysis of Experiments" 

author:
  - name: "Fall 2025"
   # affiliations:
    #  - name: "Fall 2025"
affiliation-title: "Quarter"
format: 
 pdf:
    code-fold: true
    code-line-numbers: true
    code-copy: true
    code-tools: true
    self-contained: true
    toc: false
    toc-location: left
    number-sections: true
    geometry: margin=0.2in

---
```{r setup, include=FALSE}
knitr::opts_chunk$set(message =  FALSE)
knitr::opts_chunk$set(warning =  FALSE)
knitr::opts_chunk$set(error =  FALSE)

```



:::callout
<div style="text-align: center">
<span style="color: blue;"> **STUDENT NAME **</span> 
</div>
- Alex Dieter (A070X06)
- STUDENT 2    (NetID 2)
- STUDENT 3    (NetID 3)
- STUDENT 4    (NetID 4)
:::


::: {.callout-caution}
# Due Date
**Due Date:** Monday, December 8, 2025, 11:59 PM
:::


# Introduction
Reaction time
-	Clear statement of the objective or research question.
-	Brief context or motivation.

# Experimental Design
- Description of factors and treatment structure.
- Clearly state what you are measuring and the units.
Examples: Number of words recalled (count), reaction time (seconds), taste rating (1–5 scale).
- Identify which factors are fixed vs. random.
- Description of design type (CRD, RCBD, factorial, etc.).
- Explain how randomization, replication, and (if used) blocking were implemented.
- Sample size: Provide number of observations per condition.
Guideline: 5–10 per treatment for CRD, 3–5 blocks for RCBD, total feasible within 1 hour.

(You are encouraged to explore more resources for determining the sample size )



# Data Collection
- **Procedure**: Describe how and when the experiment was conducted (e.g., location, date, steps taken).
- **Challenges/Adjustments**: Mention any difficulties or changes made during data collection (e.g., technical issues, time adjustments).
- **Data Presentation**: Display the collected data in tables or graphs, summarizing key measures like mean and standard deviation..

# Analysis

- **Exploratory Data**: Start with basic statistics (mean, SD) and visualizations (e.g., boxplots) to understand the data.


- **Hypothesis Testing**: Test your hypothesis with an appropriate statistical test (e.g., ANOVA).

- **Tables, Figures, & Code**: Include key results (ANOVA table, post-test) and relevant `R` code excerpts where needed.

-	Use `R` to analyze the data.

# Conclusions
-	Summarize key findings.
- Comment on limitations and possible improvements.


# References
 (If needed.)

# Appendices
 
 - `R` code.

	•	Factor A: Lights
Levels: off / on
	•	Factor B: Music
Levels: no / yes
	•	Factor C: Time of day
Levels: morning / evening
	•	Blocks: subject (4 people)
	•	Design type: 3-factor factorial with blocking on subject
	•	Treatments: 2 × 2 × 2 = 8 treatment combinations
	•	Experimental units: each subject will complete all 8 conditions
	•	Total experimental units: 4 subjects × 8 conditions = 32
	•	Trials per unit: 8 trials of reaction time per condition
	•	Total recorded trials: 4 × 8 × 8 = 256
	•	Randomization: randomize the order of the 8 conditions within each subject

This is equivalent to an RCBD with a 2×2×2 factorial treatment structure, exactly what is covered in PSTAT 122.

This is the only feasible and statistically valid way to include 3 factors with 4 subjects.

Setting up Randomization part
```{r}
library(dplyr)

set.seed(122)

subjects <- 1:4
lights_levels <- c("off", "on")
music_levels  <- c("no", "yes")

# full factorial 2x2
treatments <- expand.grid(
  lights = lights_levels,
  music  = music_levels
)

# function: randomize the 4 treatments for one subject
assign_order <- function(subject_id) {
  treatments %>%
    mutate(
      subject = subject_id,
      order = sample(1:4)  # randomly permuted order for this subject
    )
}

# apply to all 4 subjects
random_schedule <- bind_rows(
  assign_order(1),
  assign_order(2),
  assign_order(3),
  assign_order(4)
) %>%
  arrange(subject, order)

random_schedule
```